---
title: "NGS Project"
author: "Pietro Boscaratto, Larissa Pereira Moraes"
date: "2026-05-14"
output: pdf_document
---

# FOXG1-dependent dysregulation of GABA/glutamate neuron differentiation in autism spectrum disorders

"Summary and packages used"

```{r setup, include=FALSE, message=FALSE}
library(tidyr)
library(tidyverse)
library(readr)
library(ggplot2)
library(omicsdata)
library(limma)
library(EDASeq)
library(edgeR)
library(dplyr)
library(plyr)
library(corrplot)
library(ggfortify)
library(genefilter)
library(ggrepel)
library(pheatmap)
library(cluster)
library(knitr)
library(DESeq2)
library(clusterProfiler)
library(pathview)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ReactomePA)
library(enrichplot)
library(ggnewscale)
library(ggupset)
library(ggridges)
```

## Loading and Preprocessing

...

```{r}
data <- readRDS("/opt/omicsdata/datasets/SRP047194.rds")

# extract raw strings into data frame
raw <- data.frame(
  sample = colnames(data), 
  raw_attr = as.character(data$sra.sample_attributes)
)

# split key-value pairs for attributes and attribute values
tidy_attr <- raw %>%
  separate_rows(raw_attr, sep = "\\|") %>%
  separate(raw_attr, into = c("key", "value"),
           sep = ";;", fill = "right") %>%
  mutate(key = str_trim(key),
         value = str_trim(value))

# pivot wider so each attribute becomes its own column
# modifying last attribute to specify ASD or Non-ASD and remove redundant data
sample_meta <- tidy_attr %>%
  pivot_wider(names_from = key, values_from = value) %>%
  mutate(condition = if_else(str_detect(source_name, "non-ASD"), "non-ASD", "ASD")) %>%
  dplyr::select(-source_name, -`cell type`) %>%
  dplyr::rename(days_in_vitro = `number of days in vitro in terminal differentiation conditions`) %>%
  filter(days_in_vitro != 0)  # ---> CHECKPOINT CORRECT

# exploratory data analysis

# how many samples per condition
sample_meta <- sample_meta %>%
  mutate(days_in_vitro = as.numeric(days_in_vitro))

# Modify output to prettify
glimpse(sample_meta)
#dplyr::count(sample_meta, condition, days_in_vitro)

# filter raw count inside summarized experiment

data <- data[rowSums(assay(data, "raw_counts")) > 20, ]  

# only samples present in sample_meta
data <- data[, colnames(data) %in% sample_meta$sample]  # --> checkpoint 2

# replace sample_attributes inside data

data$sra.sample_attributes <- sample_meta
```

## Exploratory Data Analysis

...

### Sample Distribution

```{r}
ggplot(sample_meta, aes(x = condition, fill = condition)) +
  geom_bar() + 
  facet_wrap(~ days_in_vitro, labeller = label_both) + 
  labs(
    title = "Sample distribution by condition and timepoint",
    x = "Condition",
    y = "Number of sample"
  ) +
  theme_minimal()
```

### Log-transformed Boxplots

"Why we use log"

```{r}
# log-transformed
col <- factor(sample_meta$condition)
levels(col) <- c("red", "blue")
col <- as.character(col)

raw_count <- assay(data, "raw_counts")

par(mfrow = c(1, 1))

boxplot(log1p(raw_count),
        main = "Log-transformed counts",
        ylab = "log1p counts",
        las = 2,
        cex.axis = 0.7,
        col = col)

legend("topright",
       legend = c("ASD", "non-ASD"),
       fill = c("red", "blue"))
```

### RLE

"Why it sucks"

```{r}
plotRLE(raw_count,
        outline = FALSE,
        las = 2,
        ylab = "RLE",
        main = "RLE of raw counts",
        col = col,
        cex.axis = 0.5)

legend("topright",
       legend = c("ASD", "non-ASD"),
       fill = c("red", "blue"))

# RLE - check outlier samples
rle_matrix <- log1p(raw_count) - rowMedians(log1p(raw_count))
rle_medians <- colMedians(rle_matrix)

# find samples that deviate most from zero
names(rle_medians) <- colnames(raw_count)
sort(abs(rle_medians), decreasing = TRUE)

```

### PCA

"No clear distinctions"

```{r, message=FALSE}
pca <- prcomp(t(log1p(raw_count)))
summary(pca)$importance

autoplot(pca)

# by condition -> confirm paper findings
autoplot(pca, data = sample_meta, colour = "condition", 
         main = "PCA colored by condition")

autoplot(pca, data = sample_meta, colour = "days_in_vitro",
         main = "PCA colored by timepoint")
```

### MA and MD Plots

```{r}
# MA plot
limma::plotMA(log1p(raw_count),
              main = "MA plot",
              xlab = "A",
              ylab = "M")

# MD plot
MDPlot(raw_count, c(1, 7))
```

### Low Count Genes Check and Filtering

...

```{r}
# check how many genes have all zero counts
table(rowSums(raw_count) == 0)

# check how many genes have very low counts
table(rowSums(raw_count) < 10)

table(rowSums(raw_count) < 20) # Ask why so low

raw_count <- raw_count[rowSums(raw_count) >= 20, ]
```

### RLE Before Normalization

```{r}
# RLE before normalization
plotRLE(raw_count, outline = FALSE, las = 2,
        ylab = "RLE", main = "RLE of raw counts", col = col)
```

## Normalization

...

```{r, fig.show='hold', out.width="50%"}
# GC content
par(mar=c(4,4,.1,.1))
ensid <- as.character(rowRanges(data)$gene_id)
ensid <- substr(ensid, 1, 15)

gc <- getGeneLengthAndGCContent(id = ensid, org = "hg38", mode = "org.db")
table(is.na(gc[, 2]))

gc <- gc[!is.na(gc[, 2]), ]
rowData(data)$ensid <- ensid
data <- data[rowData(data)$ensid %in% rownames(gc), ]
raw_count <- assay(data, "raw_counts")
row.names(raw_count) <- row.names(gc)

set <- newSeqExpressionSet(raw_count,
                           phenoData = AnnotatedDataFrame(data.frame(conditions = factor(1:ncol(raw_count)),
                                                                     row.names = colnames(raw_count))),
                           featureData = AnnotatedDataFrame(data.frame(gc = gc[, 2], l = gc[, 1])))
#set

# Before normalization

before <- EDASeq::biasPlot(set, "gc", ylim = c(0, 10), log = TRUE, main = "Before Normalization")
lrt <- log(raw_count[, 1] + 1) - log(raw_count[, 13] + 1)

# withinlanenormalization
wt <- withinLaneNormalization(set, "gc", which = "upper", offset = TRUE)
#wt

after <- biasPlot(wt, "gc", ylim = c(0, 10), log = TRUE, main = "After Normalization") # Show them side by side
```

### Bias Boxplot

...

```{r}
biasBoxplot(lrt, gc[, 2], outline = FALSE, xlab = "GC-content", ylab = "Gene expression", las = 2, cex.axis = 0.5)
```

### Between Lane Normalization

#### Calculations

```{r}
# EDASeq between-sample
set_full <- betweenLaneNormalization(set, which = "full",  offset = TRUE)
set_uq   <- betweenLaneNormalization(set, which = "upper", offset = TRUE)
full      <- set_full@assayData$normalizedCounts
uq_edaseq <- set_uq@assayData$normalizedCounts

raw_count <- counts(set)

# edgeR

lbs <- calcNormFactors(raw_count, method = "RLE")
rle <- cpm(raw_count, lib.size = lbs * colSums(raw_count))

lbs <- calcNormFactors(raw_count, method = "TMM")
tmm <- cpm(raw_count, lib.size = lbs * colSums(raw_count))

lbs <- calcNormFactors(raw_count, method = "upperquartile")
uq  <- cpm(raw_count, lib.size = lbs * colSums(raw_count))

tc  <- cpm(raw_count)

# PLOTS EVALUATION

col <- factor(sample_meta$condition)
levels(col) <- c("red", "blue")
col <- as.character(col)
col_numeric <- as.numeric(factor(sample_meta$condition))
```

#### Boxplots

**Before Normalization**

```{r}
boxplot(log2(raw_count + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "Before normalization", xaxt = "n")
```


**Total Count Normalization**

```{r}
boxplot(log2(tc   + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "Total count", xaxt = "n")
```


**RLE Normalization**
```{r}
boxplot(log2(rle  + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "RLE", xaxt = "n")
```


**TMM Normalization**

```{r}
boxplot(log2(tmm  + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "TMM", xaxt = "n")
```

**Upper Quartile Normalization**

```{r}
boxplot(log2(uq   + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "Upper quartile", xaxt = "n")
```


**Full Quantile Normalization**
```{r}
boxplot(log2(full + 1), outline = FALSE, las = 2, col=col,
        ylab = "log2(count+1)", main = "Full quantile", xaxt = "n")
```

#### RLE plots

```{r}
EDASeq::plotRLE(raw_count, outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "Before normalization", xaxt = "n")
```


**Total Count**

```{r}
EDASeq::plotRLE(tc,   outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "Total count",          xaxt = "n")
```


**RLE**

```{r}
EDASeq::plotRLE(rle,  outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "RLE",                  xaxt = "n")
```

**TMM**

```{r}
EDASeq::plotRLE(tmm,  outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "TMM",                  xaxt = "n")
```

**Upper Quartile**

```{r}
EDASeq::plotRLE(uq,   outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "Upper quartile",       xaxt = "n")
```

**Full Quantile**

```{r}
EDASeq::plotRLE(full, outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "Full quantile", xaxt = "n")
```

### Correlation Plots

```{r}
tmm_colnames <- colnames(tmm)
colnames(tmm) <- substr(colnames(tmm), 1, 8)
m <- cor(log1p(tmm))
corrplot(m, method = "color", order = "hclust", tl.cex = 0.2,
         main = "Correlation - TMM")
colnames(tmm) <- tmm_colnames
```

### PCA After Normalization

```{r}
asd_pca <- prcomp(t(log1p(tmm)))
#summary(asd_pca)$importance[, 1:10]

autoplot(asd_pca, data = sample_meta, colour = "condition",
         main = "PCA - TMM, coloured by condition")
autoplot(asd_pca, data = sample_meta, colour = "days_in_vitro",
         main = "PCA - TMM, coloured by days in vitro")
```

### PCA with regard to different normalization

"Check if it's relevant, make it pretty"

```{r}
EDASeq::plotPCA(tc,   k = 3, labels = FALSE, col = col_numeric, pch = 20,
                main = "PCA - Total count")
EDASeq::plotPCA(tmm,  k = 3, labels = FALSE, col = col_numeric, pch = 20,
                main = "PCA - TMM")
EDASeq::plotPCA(rle,  k = 3, labels = FALSE, col = col_numeric, pch = 20,
                main = "PCA - RLE")
EDASeq::plotPCA(full, k = 3, labels = FALSE, col = col_numeric, pch = 20,
                main = "PCA - Full quantile")
```

## Differential Gene Expression

"biological question is to identify the transcriptional differences between ASD vs non-ASD at days 11 vs days 31"

### Mean-Variance

```{r}
data$sra.sample_attributes <- dplyr::select(data$sra.sample_attributes, days_in_vitro) |>
  unlist() |>
  as.character()

design <- model.matrix(~sra.sample_attributes, data = colData(data))
head(design)

dge <- DGEList(tmm) # use normalized raw count -rle same results

dge <- calcNormFactors(dge, method="TMM")
dge <- estimateDisp(dge, design)

plotMeanVar(dge, show.raw.vars = TRUE, show.ave.raw.vars = F, NBline = TRUE, show.tagwise.vars = TRUE) # Ask why it doesn't fit the grey points
```

### Differentially expressed genes

```{r, message=FALSE}
# Identify differentially expressed genes
fit <- glmFit(dge, design)
#fit$offset[,1:20]

res <- glmLRT(fit, coef = 2)
topTags(res, sort.by = "logFC")

tab <- topTags(res, n=Inf)
tab <- tab$table
head(tab)

de <- tab[tab$FDR <0.05,]

plotMD(res)
```


#### Volcano Plots

```{r}
# volcano plots
names <- row.names(de)

tab$sign <- NA
tab$sign[tab$logFC < 0] <- "Down"
tab$sign[tab$logFC > 0] <- "Up"
tab$sign[tab$FDR >= 0.05] <- "NoSig"
tab$gene <- rownames(tab)

ggplot(tab, aes(x = logFC, y = -log10(PValue), color = sign)) + geom_point() + theme_classic() +
  scale_color_manual(values = c("Down" = "blue", "Up" = "red", "NoSig" = "black")) +
  geom_text_repel(data = tab[1:20, ], aes(label = gene), size = 2.5, max.overlaps = 20) 
```

#### P-values plot

```{r}
hist(tab$PValue, main="p-values distribution")
```

#### Cluster Analysis

```{r}
# cluster analysis

data_degs <- tmm[names, ]


Col <- data.frame(Subtype=data$sra.sample_attributes)
rownames(Col) <- colnames(data)

ann_colors = list(Subtype = c("11"="blue","31"="lightblue"))

g <- pheatmap(log1p(data_degs), show_colnames = FALSE, scale = "row", show_rownames = FALSE,
              drop_levels = TRUE, annotation_colors = ann_colors, annotation_col = Col)

# silhouette index
cut <- cutree(g$tree_col, k = 2)

dist_mat <- dist(t(log1p(data_degs)))
plot(silhouette(cut, dist_mat), col=c("#2E86C1", "#EC7063"), border=NA)

```


## Differential Expression Gene Analysis

```{r}
# Rank differentially expressed genes

top <- topTags(res, n = Inf)$table

# significant genes
deg <- top[top$FDR <= 0.05,]
dim(deg)


# extract gene ids
universo <- substr(row.names(top), 1, 15)
sign <- substr(row.names(deg), 1, 15)

# convert ensembl to entrez id
universo.entrez <- mapIds(org.Hs.eg.db,
                          keys = universo,
                          column = "ENTREZID",
                          keytype = "ENSEMBL")
sign.entrez <- mapIds(org.Hs.eg.db,
                      keys = sign, #Column containing Ensembl gene ids
                      column = "ENTREZID",
                      keytype = "ENSEMBL")
universo.entrez <- as.character(universo.entrez)
sign.entrez <- as.character(sign.entrez)
```


### KEGG Enrichment Analysis

```{r}
kk <- enrichKEGG(
  gene = sign.entrez,
  universe = universo.entrez,
  organism = "hsa"
)

# kegg visualization
barplot(kk) # Quickly rank terms by the number of genes involved
dotplot(kk) # Useful when you want to distinguish between a term that is significant but contains few of your genes vs. one where your DEGs dominate the pathway.
cnetplot(kk) # Useful to see gene overlap between terms and identify multi-pathway genes
heatplot(kk, showCategory = 10) #Reveals which genes are shared across multiple terms and which are unique to one pathway

```


### Pathway Similarity Analysis


```{r}
edo <- pairwise_termsim(kk)
emapplot(edo)
upsetplot(edo, 8)


ego <- enrichGO(gene = sign.entrez,
                universe = universo.entrez,
                OrgDb = 'org.Hs.eg.db',
                keyType = "ENTREZID",
                ont = "BP",
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                readable = TRUE)

View(ego@result[1:10,])

barplot(ego)
upsetplot(ego, 5)
```


### GSEA Analysis

```{r}
tt <- top$logFC
names(tt) <- universo.entrez

table(duplicated(names(tt)))
tt <- tt[!duplicated(names(tt))]

sort_tt <- sort(tt, decreasing = TRUE)

kk2 <- gseKEGG(
  geneList = sort_tt,
  organism = "hsa",
  pvalueCutoff = 0.05
  
)

View(kk2@result[, 1:7])


kkGO <- gseGO(
  geneList = sort_tt,
  ont = "BP",
  OrgDb = 'org.Hs.eg.db',
  keyType = "ENTREZID",
  pAdjustMethod = "BH"
)

View(kkGO@result[1:10,])

```

